Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-16 ICML-submission/algorithm-for-strongly-linearly-separable-data.tex
thm 12 for randomized

Define a filtration $\cbr{\calB_j}_{j=0}^M$,
where $\calB_j = \sigma((x_1, y_1), \ldots, (x_{(K-1)j}, y_{(K-1)j}))$
for every $j$ in $\{1,2,\dots,M\}$.
By Claim 2 of~\citet{Daniely-Helbertal-2013}, as $\ell_j$
is chosen uniformly from $\cbr{1,\dots,K}$ and independent of $\calB_{j-1}$,
$$
\Exp \sbr{ \sum_{t=(K-1)(j-1) + 1}^{(K-1)j} z_t ~\Bigg|~ \calB_{j-1} } \geq \frac{K-1}2 \; .
$$
This implies that
$$
\Exp \sbr{ \sum_{t=(K-1)(j-1) + 1}^{(K-1)j} z_t } \ge \frac{K-1}2 \; .
$$
Summing over all $j$ in $\{1,2,\dots,M\}$,
$$
\Exp \sbr{ \sum_{t=1}^{(K-1)M} z_t} \geq \frac{K-1}2 \cdot M = \frac{K-1}2 \left\lfloor \frac 1 4 (R/\gamma)^2 \right\rfloor \; .
$$

>>>>

Define a filtration $\cbr{\calB_j}_{j=0}^M$,
where $\calB_j = \sigma((x_1, y_1, \hat{y}_1), \ldots, (x_{(K-1)j}, y_{(K-1)j}, \hat{y}_{(K-1)j}))$
for every $j$ in $\{1,2,\dots,M\}$.
By Claim 2 of~\citet{Daniely-Helbertal-2013}, as $\ell_j$
is chosen uniformly from $\cbr{1,\dots,K}$ and independent of $\calB_{j-1}$ and $\calA$'s
randomness,
$$
\Exp \sbr{ \sum_{t=(K-1)(j-1) + 1}^{(K-1)j} z_t ~\Bigg|~ \calB_{j-1} } \geq \frac{K-1}2 \; .
$$
This implies that
$$
\Exp \sbr{ \sum_{t=(K-1)(j-1) + 1}^{(K-1)j} z_t } \ge \frac{K-1}2 \; .
$$
Summing over all $j$ in $\{1,2,\dots,M\}$,
$$
\Exp \sbr{ \sum_{t=1}^{(K-1)M} z_t} \geq \frac{K-1}2 \cdot M = \frac{K-1}2 \left\lfloor \frac 1 4 (R/\gamma)^2 \right\rfloor \; .
$$

### 2019-01-16 ICML-submission/appendix-multiclass-perceptron.tex
pass over intro
thm 12 for randomized

\begin{algorithm}[h]
\caption{\textsc{Online Multiclass Classification}
\label{algorithm:mutliclass-classification}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Predict $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:online-multiclass-classification-mistake-lower-bound}
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Online Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are weakly linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

\begin{proof}
Let $T = \lfloor (R/\gamma)^2 \rfloor$ and let $x_1, x_2, \dots, x_T$ be the
orthogonal vectors such that $\norm{x_t} = R$ for all $t=1,2,\dots,T$. For
example, we can take $V = \R^T$ and $x_t = R e_t$ where $e_t$ is $t$-th element
of the standard orthonormal basis of $\R^T$.

Since the algorithm is deterministic, we can construct the sequence of labels
$y_1, y_2, \dots, y_T$ adaptively based on the predictions $\widehat y_1,
\widehat y_2, \dots, \widehat y_T$ of the algorithm. We define $y_t$ to be any
element of $\{1,2,\dots,K\}$ not equal to $\widehat y_t$. This way the algorithm
makes a mistake in every round $t=1,2,\dots,T$.

It remains to show that the examples we have constructed are linearly separable
with margin $\gamma$. To prove that we demonstrate vectors $w_1, w_2, \dots, w_K$
satisfying conditions \eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2}. We define
$$
w_i = \frac{1}{R\sqrt{T}} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} x_t \qquad \text{for $i=1,2,\dots,K$.}
$$
Let $a_i = |\{ t ~:~ 1 \le t \le T, \ y_t = i \}|$ be the number of occurrences of label $i$.
It is easy to see that
$$
\norm{w_i}^2 = \frac{1}{R^2 T} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} \norm{x_t}^2 = \frac{a_i}{T} \qquad \text{for $i=1,2,\dots,K$.}
$$
Since $\sum_{i=1}^K a_i = T$, the condition
\eqref{equation:weak-linear-separability-1} holds. To verify condition
\eqref{equation:weak-linear-separability-2} consider any labeled example $(x_t,
y_t)$. Then, by definition of $w_{y_t}$ we have
\begin{align*}
\ip{w_{y_t}}{x_t}
& = \frac{1}{R\sqrt{T}} \sum_{\substack{s : 1 \le s \le T \\ y_s = y_t}} \ip{x_s}{x_t} \\
& = \frac{1}{R\sqrt{T}} \norm{x_t}^2 \\
& = \frac{R}{\sqrt{T}} \\
& \ge \frac{R}{R/\gamma} \\
& = \gamma \; .
\end{align*}
and for any $i \in \{1,2,\dots,K\} \setminus \{y_t\}$ we have
$\ip{w_i}{x_t} = 0$. Therefore, condition \eqref{equation:weak-linear-separability-2} holds.
\end{proof}

>>>>

\begin{protocol}[h]
\caption{\textsc{Online Multiclass Classification}
\label{algorithm:mutliclass-classification}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Predict $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\ENDFOR
}
\end{algorithmic}
\end{protocol}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:online-multiclass-classification-mistake-lower-bound}
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any (possibly randomized) algorithm
$\calA$ for the
\textsc{Online Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are weakly linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\frac 1 2 \lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

\begin{proof}
Let $T = \lfloor (R/\gamma)^2 \rfloor$, $V = \R^T$, and for all $t$ in
$\cbr{1,\ldots,T}$, define instance $x_t = R e_t$ where $e_t$ is $t$-th element
of the standard orthonormal basis of $\R^T$.
Let labels $y_1, \ldots, y_T$ be chosen i.i.d uniformly at random from
$\cbr{1,2,\ldots,K}$ and independently of any randomness used by the algirithm
$\calA$.

%let $x_1, x_2, \dots, x_T$ be the
%orthogonal vectors such that $\norm{x_t} = R$ for all $t=1,2,\dots,T$. For
%example, we can take

%Since the algorithm is deterministic, we can construct the sequence of labels
%$y_1, y_2, \dots, y_T$ adaptively based on the predictions $\widehat y_1,
%\widehat y_2, \dots, \widehat y_T$ of the algorithm. We define $y_t$ to be any
%element of $\{1,2,\dots,K\}$ not equal to $\widehat y_t$. This way the algorithm
%makes a mistake in every round $t=1,2,\dots,T$.

We first show that the set of examples $(x_1,y_1)$, $\ldots$,
$(x_T, y_T)$ we have constructed is weakly linearly separable
with margin $\gamma$. To prove that, we demonstrate vectors $w_1, w_2, \dots, w_K$
satisfying conditions \eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2}. We define
$$
w_i = \frac{\gamma}{R} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} e_t \qquad \text{for $i=1,2,\dots,K$.}
$$
Let $a_i = |\{ t ~:~ 1 \le t \le T, \ y_t = i \}|$ be the number of occurrences of label $i$.
It is easy to see that
$$
\norm{w_i}^2 = \frac{\gamma^2}{R^2} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} \norm{e_t}^2 = \frac{a_i \gamma^2}{R^2} \qquad \text{for $i=1,2,\dots,K$.}
$$
Since $\sum_{i=1}^K a_i = T$,
$\sum_{i=1}^K \norm{w_i}^2 = T \cdot \frac{\gamma^2}{R^2} \leq 1$, i.e.
the condition
\eqref{equation:weak-linear-separability-1} holds. To verify condition
\eqref{equation:weak-linear-separability-2} consider any labeled example $(x_t,
y_t)$. Then, for any $i$ in $\cbr{1,\ldots,K}$, by the definition of $w_i$, we have
\begin{align*}
\ip{w_i}{x_t}
& = \frac{\gamma}{R} \sum_{\substack{s : 1 \le s \le T \\ y_s = i}} \ip{e_s}{R e_t} \\
& = \gamma \cdot \sum_{\substack{s : 1 \le s \le T \\ y_s = i}} \one[s = t] \\
& = \gamma \cdot \one[y_t = i]
\; .
\end{align*}
Therefore, if $i = y_t$, $\ip{w_i}{x_t} = \gamma$;
otherwise $i \neq y_t$, in which case $\ip{w_i}{x_t} = 0$.
Hence, condition \eqref{equation:weak-linear-separability-2} holds.

We now give a lower bound on the number of mistakes $\calA$ makes.
Define a filtration $\cbr{\calB_t}_{t=0}^T$,
where $\calB_t = \sigma((x_1, y_1, \hat{y}_1), \ldots, (x_t, y_t, \hat{y}_t))$.
As $y_t$ is chosen uniformly from $\cbr{1,\ldots,K}$ and independent of
$\calB_{t-1}$ and $\hat{y}_t$,
\[ \Exp[ \indicator{\widehat y_t \neq y_t} | \calB_{t-1} ] \geq 1 - \frac 1 K \geq \frac 1 2. \]
This implies that
\[ \Exp[ \indicator{\widehat y_t \neq y_t} ] \geq \frac 1 2. \]
Summing over all $t$ in $\cbr{1,\ldots,T}$, we conclude that
\[ \Exp \sbr{ \sum_{t=1}^T \indicator{\widehat y_t \neq y_t} } \geq \frac T 2 = \frac 1 2 \lfloor (R/\gamma)^2 \rfloor, \]
which completes the proof.
\end{proof}

%& = \frac{R}{\sqrt{T}} \\
%& \ge \frac{R}{R/\gamma} \\
%& = \gamma \; .
%and similarly, for any $i \in \{1,2,\dots,K\} \setminus \{y_t\}$ we have
%$\ip{w_i}{x_t} = 0$.

### 2019-01-16 ICML-submission/notions-of-linear-separability.tex
before editing sect. 5
thm 12 for randomized

The notion of weak linear separability with a margin is standard. It is used in
the full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. It is a folklore result that if a set
of labeled examples is separable with a margin $\gamma$ and the norm of the
examples is bounded by $R$, then \textsc{Multiclass Perceptron} algorithm makes
at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore
result is that \textsc{Multiclass Perceptron} is essentially optimal in the
sense that any deterministic algorithm must make $\left\lfloor (R/\gamma)^2
\right \rfloor$ mistakes in the worst case.
\chicheng{And any randomized algorithm must make $\left\lfloor (R/\gamma)^2
\right \rfloor / 2$ mistakes?}
For completeness, we state the
\textsc{Multiclass Perceptron} algorithm and give proofs of both of these
results in Appendix~\ref{section:multiclass-perceptron-proofs}.

\chicheng{Shall we mention the multiclass margin notion that appeared in e.g.}

>>>>

The notion of weak linear separability with a margin is standard. It is used in
the full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. It is a folklore result that if a set
of labeled examples is separable with a margin $\gamma$ and the norm of the
examples is bounded by $R$, then \textsc{Multiclass Perceptron} algorithm makes
at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore
result is that \textsc{Multiclass Perceptron} is essentially optimal in the
sense that any deterministic algorithm must make $\frac 1 2 \left\lfloor (R/\gamma)^2
\right \rfloor$ mistakes in the worst case.
For completeness, we state the
\textsc{Multiclass Perceptron} algorithm and give proofs of both of these
results in Appendix~\ref{section:multiclass-perceptron-proofs}.

\chicheng{Shall we mention that this multiclass margin notion has also appeared in e.g.~\cite{Crammer-Singer-2003}?}
